A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-07 paper.tex
Minor text changes

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

@@

\section*{Ancestral graphs and stochastic processes}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
recombination
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
\cite{kelleher2016efficient} provide a description of his algorithm along with
an efficient implementation, the details of which turn out to be highly relevant in 
later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

The Ancestral Recombination Graph (ARG) was later introduced by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues used the word
``ARG'' to refer to both a specific branching-coalescing stochastic process and
the graph structure derived from this process. For clarity, we will refer to the
process as the Griffiths process, or the coalescent with recombination (CwR)
\citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure} if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities (see the ``\nameref{probabilities}'' section).

[Review of other work done on the ARG stochastic process here.
Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
clarified the relationship between Hudson's algorithm and the ARG.
Explain that ARG simulation is inherently less efficient than Hudson's
algorithm because we must track all lineages, even those that correspond
to sections that have coalesced, or we won't have a complete ARG.
This might be tricky to communicated without a figure, but is an
esoteric point perhaps not worth spending a figure on.]

@@

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
% FIXME: need some proper statistical language here to describe why these
% methods didn't work so well.
These methods met with limited success
due to the overwhelmingly large state space of ARGs, which also has
no simple geometry or neighbourhood structure that an inference or
sampling method could exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping}
became important at this point,
since parsimonious ARGs have no immediate connection to
the stochastic process. Indeed, the ARG instances produced by
the coalescent with recombination are usually highly unparsimonious,
since they contain many invisible recombinations
(although those produced under the SMC may be parsimonious, depending
on the pattern of mutations that is present).

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination,
but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
The fundamental units of an EARG are the \emph{events} that correspond directly
to those that
occur in the ARG stochastic process described in the previous section.
Each event corresponds to a node in the graph, and there are therefore
two types of node: common ancestor and recombination. As in the
stochastic process, edges represent lineages, and those lineages
are modified by the events that they pass through.

@@

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(Note that this formulation is equivalent to stipulating two
different ``types'' of node and only storing breakpoints
with recombination nodes).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The Griffiths ARG encoding is also limited in the types of ancestral
history that can be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough that,
with high probability, the number of concurrent lineages in the ARG remains
much smaller than $N$ at each time. While these approximations are
reasonable
to make when studying the Kreitman \emph{Drosphila Melanogaster}
dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
at 43 variant sites over x Kb \textcolor{red}{details and citation
for Ne estimate}) that is classically used for benchmarking
ARG inference methods [cites],
they are highly questionable in modern datasets. % absurd really
For example, there are now many
human datasets consisting of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

Another limitation of the Griffiths ARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

>>>>

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
% Do we mean "despite the fact that analysis of genetic data relies on methods
% which ultimately derive from the inheritance paths defined by genetic trees
% and the mutations upon them, analysis so far has focussed on summaries ...
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

@@

\section*{Ancestral graphs and stochastic processes}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
recombination
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
\cite{kelleher2016efficient} provide a description of his algorithm along with
an efficient implementation, the details of which turn out to be highly relevant to 
later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

Although Hudson's algorithm was accompanied by a graph
schematic~\citep[][Fig. 5]{hudson1983properties}, the term
``Ancestral Recombination Graph'' or ARG, was introduced later by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral, griffiths1997ancestral},
who provided a closely related and complementary treatment.
Where Hudson's algorithm focusses on efficient \emph{simulation}, the original
ARG literature is focused on mathematical results about the underlying stochastic
process. Griffiths and colleagues used the word
``ARG'' to refer to both a specific branching-coalescing stochastic process and
the graph structure derived from this process, but for clarity and following recent usage,
we will use ARG to refer to a graph structure, and refer to the
process as the Griffiths process or the coalescent with recombination
\citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure} if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph than the Hudson algorithm (see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

@@

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?

% NB: should this para go into the previous section? It would be
% clean to start the section with the subsequent para
% starting "The Griffiths ARG encoding is an elegant and mathematically economical description of the CwR"
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
% FIXME: need some proper statistical language here to describe why these
% methods didn't work so well.
These methods met with limited success
due to the overwhelmingly large state space of ARGs, which also has
no simple geometry or neighbourhood structure that an inference or
sampling method could exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
This highlighted the distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
since parsimonious ARGs have no immediate connection to
the stochastic process. Indeed, the ARG instances produced by
the coalescent with recombination are usually highly unparsimonious,
since they contain many invisible recombinations
% Do we need the following caveat?
(although those produced under the SMC may be parsimonious, depending
on the pattern of mutations that is present).

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination, whose
fundamental units are the  \emph{events} that correspond directly
to those that occur in the associated stochastic process. We will refer
to this ARG encoding as an ``Event ARG'', or eARG: 
an example is given in Fig.~\ref{fig-arg-data-structure}.
Each node in the graph corresponds to an event: therefore the graph
contains both recombination nodes (node \textsf{D}) and common ancestor
nodes (nodes \textsf{E}, \textsf{F}, and \textsf{G}) corresponding to their
respective event types. Edges represent lineages,
and those lineages are modified by the events that they pass through.

@@

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
a recombination node is of a different ``type'' to other nodes,
with breakpoints only associated with this specific node type).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough that,
with high probability, the number of concurrent lineages in the ARG remains
much smaller than $N$ at each time. While these approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
as used in the ``\nameref{ARG_inference_methods}'' section),
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

A final limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure} is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

% Potentially useful text from a previous iteration:
% Wiuf and Hein's classical pair of papers
% clarified the relationship between Hudson's algorithm and the ARG.
% ARG simulation is inherently less efficient than Hudson's
% algorithm because we must track all lineages, even those that correspond
% to sections that have coalesced, or we won't have a complete ARG.
